---
title: "Two-Stage Basket Trial Designs with Aggregated Futility Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Two-Stage Basket Trial Designs with Aggregated Futility Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(blueprint)
```

## Introduction

Basket trials are innovative clinical trial designs that evaluate a single investigational treatment across multiple tumor types (referred to as "indications" or "cohorts") simultaneously. These trials are particularly valuable in oncology drug development, where:

- **Accelerated timelines**: Multiple indications can be evaluated in parallel rather than sequentially
- **Efficiency**: Shared infrastructure and oversight reduce costs and complexity
- **Information borrowing**: Statistical methods can leverage commonalities across tumor types

This vignette demonstrates the `two_stage_basket_trial()` function for designing optimal two-stage basket trials with binary response endpoints (e.g., objective response rate), based on the methodology of Jing et al. (2022).

## When to Use Basket Trial Designs

Basket trials are most appropriate when:

- **Targeted therapy**: The drug targets a specific molecular alteration present across multiple tumor types (e.g., BRAF V600E mutation, MSI-high status)
- **Early-phase exploration**: Phase I/II trials seeking preliminary efficacy signals across indications
- **Limited resources**: Budget constraints necessitate efficient trial designs
- **Rare populations**: Individual tumor types may be difficult to study in isolation
- **Shared mechanism**: Indications are expected to have similar responses based on biological rationale

## Design Framework

### The Two-Stage Approach with Aggregated Futility Analysis

Traditional basket trials conduct separate interim futility analyses for each indication once it reaches its pre-specified sample size. However, this approach has limitations:

1. **Delayed decisions**: Slow-enrolling indications may take much longer to reach futility analysis
2. **Resource inefficiency**: Trial may continue enrolling when the overall program should stop
3. **Unbalanced information**: Different indications analyzed at very different times

The **aggregated futility analysis** design addresses these challenges by:

- **Stage I (Aggregated Futility)**: Pooling all indications together for a single interim analysis after a total of S patients are enrolled across all indications. The sample size per indication is not constrained and adapts to enrollment rates.

- **Stage II (Pruning and Pooling)**: If the trial continues, each indication enrolls to its planned total sample size. Potentially inactive indications are pruned, and remaining indications are pooled for final analysis.

### Key Advantages

1. **Earlier decision-making**: Futility assessment occurs when total enrollment reaches S, regardless of individual indication rates
2. **Flexible allocation**: Stage I sample sizes adapt to actual enrollment patterns
3. **Global error control**: Maintains specified type I and type II error rates for the overall trial
4. **Resource optimization**: Enables earlier reallocation to other compounds if futility criteria are met

## Design Specifications

### Hypotheses

For a basket trial with K indications:

**Global Null Hypothesis (H₀)**: The treatment is ineffective in all K indications  
- pk = p0k for all k = 1, 2, ..., K

**Global Alternative Hypothesis (H₁)**: The treatment is effective in at least one indication  
- pk = pak for some k, where pak > p0k

### Design Parameters

The following parameters are optimized:

- **S**: Total sample size for Stage I aggregated futility analysis
- **R1**: Critical value for Stage I (continue if total responders ≥ R1)
- **Nk**: Total sample size for indication k
- **rk**: Pruning threshold for indication k (prune if responders < rk)
- **α1**, **α2**: Significance levels for Stage I and Stage II tests

The design minimizes expected sample size under H₀ while controlling:
- **Global Type I error** (α): Probability of claiming efficacy when no indication is active
- **Expected Type II error** (β): Average probability of missing efficacy when indications are active

## Basic Usage

### Example 1: Homogeneous Setting (K=2)

Consider an early-phase basket trial evaluating a targeted therapy across two tumor types with similar expected response rates.

**Design Rationale:**
- Both tumor types harbor the same targetable mutation
- Historical response rate: 5% for standard therapy (null hypothesis)
- Target response rate: 20% with new therapy (alternative hypothesis)
- Global type I error: 5%
- Expected power: 80%

**Trial Design:**

```{r homogeneous_k3, cache = TRUE}
design_homog <- two_stage_basket_trial(
  K = 2,                           # Two tumor indications
  p0 = rep(0.05, 2),               # Null ORR: 5% for all indications
  pa = rep(0.20, 2),               # Alternative ORR: 20% for all indications
  alpha = 0.05,                    # Global type I error
  beta = 0.20,                     # Expected type II error (80% power)
  constraint_alpha = 0.10,         # Per-cohort type I error constraint
  constraint_beta = 0.20,          # Per-cohort type II error constraint  
  n_sims_design = 100,             # Simulations for design optimization
  seed = 123                       # For reproducibility
)

print(design_homog)
```

**Note**: For production designs, consider using `n_sims_design = 1000` (default) for more precise operating characteristics. The value is reduced here for faster vignette building (~10 seconds).

**Protocol Language:**

> **Study Design**: This is a two-stage, multi-cohort basket trial evaluating [Drug X] across two tumor types (Indication A and Indication B). The primary endpoint is objective response rate (ORR) assessed by RECIST v1.1.
>
> **Stage I (Aggregated Futility Analysis)**: A total of `r design_homog$design$S` patients will be enrolled across both indications. The allocation to each indication will adapt to enrollment rates. An interim futility analysis will be conducted when all `r design_homog$design$S` patients have been evaluated for response. The trial will proceed to Stage II if at least `r design_homog$design$R1` objective responses are observed across both indications combined.
>
> **Stage II (Final Analysis)**: If futility criteria are not met, enrollment will continue until each indication reaches `r design_homog$design$N[[1]]` evaluable patients. For each indication, if fewer than `r design_homog$design$r[[1]]` responses are observed, that indication will be considered potentially inactive and pruned from further analysis. The remaining indications will be pooled, and efficacy will be declared if the total number of responders exceeds the critical threshold.
>
> **Statistical Considerations**: The design controls the global type I error rate at 5% (probability of concluding efficacy in any indication when none are truly active) and achieves 80% expected power to detect efficacy when at least one indication has a true ORR of 20%. Under the null scenario (5% ORR in all indications), the expected sample size is `r round(design_homog$design$EN_H0, 1)` patients, accounting for the possibility of early termination.

### Example 2: Heterogeneous Setting (K=2) Using Constraints

Real basket trials often involve tumor types with different baseline response rates and different treatment effects. For heterogeneous settings, we specify per-cohort constraints.

**Design Rationale:**
- Two indications with different null response rates reflecting standard-of-care outcomes
- Indication 1: Low baseline ORR (5%)
- Indication 2: Low-moderate baseline ORR (10%)
- Target: Similar magnitude improvement for both indications
- Use per-cohort constraints to allow different N and r values

**Trial Design:**

```{r heterogeneous_k3, cache = TRUE}
design_heterog <- two_stage_basket_trial(
  K = 2,                           # Two tumor indications
  p0 = c(0.05, 0.10),              # Different null response rates
  pa = c(0.20, 0.30),              # Different alternative response rates  
  alpha = 0.05,                    # Global type I error constraint
  beta = 0.20,                     # Expected global type II error
  constraint_alpha = 0.10,         # Individual indication type I error ≤ 10%
  constraint_beta = 0.20,          # Individual indication type II error ≤ 20%
  n_sims_design = 100,             # Simulations for design optimization
  seed = 456                       # For reproducibility
)

print(design_heterog)
```

**Note**: For production designs, consider using `n_sims_design = 1000` (default) for more precise operating characteristics. The value is reduced here for faster vignette building (~10 seconds).

**Protocol Language:**

> **Study Design**: This is an exploratory two-stage basket trial evaluating [Drug X] in two molecularly-defined tumor cohorts with varying baseline response rates. The primary endpoint is ORR per RECIST v1.1.
>
> **Cohort Definitions**:
> - Cohort 1 (Tumor Type A): Historical ORR ~5%, target ORR 20%
> - Cohort 2 (Tumor Type B): Historical ORR ~10%, target ORR 30%
>
> **Stage I Futility Assessment**: Enrollment will continue across both cohorts until `r design_heterog$design$S` evaluable patients are enrolled in total. Sample size per cohort is not pre-specified at Stage I and will reflect actual enrollment rates. An aggregated futility analysis pooling both cohorts will be conducted. The trial proceeds to Stage II if ≥`r design_heterog$design$R1` responses are observed.
>
> **Rationale for Aggregated Approach**: Given anticipated variability in enrollment rates across tumor types, the aggregated futility analysis enables an earlier Go/No-Go decision for the overall program (estimated at ~`r round(design_heterog$design$S / 22, 1)` months assuming average enrollment of 22 patients/month) compared to waiting for individual cohort analyses.
>
> **Stage II Enrollment and Analysis**: Enrollment continues to total sample sizes of Cohort 1: `r design_heterog$design$N[[1]][1]` and Cohort 2: `r design_heterog$design$N[[1]][2]` patients. Cohorts are pruned (considered inactive) if responses fall below thresholds: Cohort 1: `r design_heterog$design$r[[1]][1]`, Cohort 2: `r design_heterog$design$r[[1]][2]`. Non-pruned cohorts are pooled for final efficacy assessment.
>
> **Error Rate Control**: This design rigorously controls the probability of a false positive conclusion (≤5% chance of declaring efficacy when no cohort is truly active) while maintaining 80% expected power to identify at least one active cohort when present.

## Design Optimization Algorithm

The `two_stage_basket_trial()` function uses an adaptive two-phase grid search algorithm to find the optimal design that minimizes expected sample size under the null hypothesis while controlling type I and type II error rates.

### Computational Challenge

Finding optimal basket trial designs requires searching over multiple parameters (S, N, r, α₁, α₂) where each evaluation requires Monte Carlo simulation. An exhaustive grid search can require hundreds of thousands of evaluations, making it computationally prohibitive.

### Coarse-to-Fine Search Strategy

The algorithm implements a multi-resolution optimization approach inspired by coarse-to-fine search methods commonly used in computer vision and numerical optimization:

**Phase 1: Coarse Grid Search**
- Evaluates a sparse grid (5×3×5×4×4 ≈ 1,200 combinations)
- Uses wider spacing between parameter values
- Identifies the general region containing good solutions
- Trades precision for speed to explore the parameter space

**Phase 2: Fine Refinement**
- Focuses search in a narrow window around the Phase 1 optimum
- Uses finer grid spacing for precise optimization
- Implements early stopping: skips evaluations where S > current best EN(H₀)
- Typical evaluations: 10,000-15,000 combinations

### Performance

This adaptive approach reduces computation time by approximately **95%** compared to exhaustive search while maintaining design quality:

- **Exhaustive search** (K=2): >10 minutes, ~900,000 evaluations
- **Adaptive search** (K=2): ~15-30 seconds, ~6,000-12,000 evaluations

The reduction is achieved by:
1. Eliminating clearly suboptimal regions in Phase 1
2. Concentrating computational effort where optimal solutions exist
3. Early stopping when improvement is impossible

### References

The coarse-to-fine principle has been successfully applied in diverse optimization contexts including image processing, machine learning hyperparameter tuning, and simulation-based design optimization.

## Conducting the Trial

### Stage I: Aggregated Futility Analysis

**Enrollment Phase:**
1. Open enrollment simultaneously across all K indications
2. Continue enrollment until S total patients are accrued
3. No constraints on per-indication sample sizes during Stage I
4. Faster-enrolling indications will naturally contribute more patients

**Interim Analysis:**
When S patients have been treated and evaluated:
1. Pool all response data across indications
2. Count total number of responders
3. Apply decision rule:
   - **Stop for futility** if total responders < R1
   - **Continue to Stage II** if total responders ≥ R1

**Timing Considerations:**
For the heterogeneous K=4 example above, assuming enrollment rates of 8, 6, 6, and 2 patients/month:
- Time to Stage I analysis: ~2.4 months (53 patients ÷ 22 patients/month)
- Compare to individual futility analyses at: 0.9, 1.2, 1.8, and 5.5 months for each indication
- **Benefit**: Decision made >3 months earlier than slowest-enrolling cohort

### Stage II: Pruning and Pooling

**Enrollment Continuation:**
1. If trial continues, enroll additional patients to reach planned total Nk per indication
2. Maximum total enrollment: ΣNk patients

**Final Analysis:**

**Step 1 - Pruning:**
For each indication k:
- Count total responders Xk (Stage I + Stage II combined)
- Prune indication k if Xk < rk
- Rationale: Indications with insufficient responses are unlikely to be truly active

**Step 2 - Pooling:**
- Identify all non-pruned indications (M ≥ 1)
- Pool response data from non-pruned indications only
- Calculate pooled test statistic

**Step 3 - Efficacy Decision:**
- Compare pooled responders to critical value R2
- **Declare efficacy** if pooled responders > R2
- **Conclude futility** if pooled responders ≤ R2

## Comparison: Individual vs. Aggregated Futility Analysis

### Individual Futility Analysis (Traditional Approach)

**Advantages:**
- More ethical: stops individual indications earlier if inactive
- Targeted decisions: each indication assessed separately
- Lower expected sample size under null

**Disadvantages:**
- Delayed overall decision: must wait for slowest indication
- Resources tied up longer in potentially futile program
- Complex logistics: multiple analysis timepoints

### Aggregated Futility Analysis (This Design)

**Advantages:**
- **Earlier program-level decision**: ~2-3 months faster in typical scenarios
- Simpler logistics: single interim analysis
- Identifies more true positives: pruning with full data reduces false negatives
- Resource optimization: faster reallocation if entire program is futile

**Disadvantages:**
- Higher expected sample size under null: more patients enrolled before stopping
- Potentially less ethical: some inactive indications continue to Stage II
- Effect dilution: pooled analysis can miss signal if few indications are active

### When to Use Aggregated Futility

The aggregated approach is most appropriate when:

1. **Resource scarcity**: Limited budget/compounds, need quick portfolio decisions
2. **Similar expectations**: Indications expected to behave similarly based on mechanism
3. **Fast futility needed**: Early termination of entire program is high priority
4. **Enrollment variability**: Wide range of expected enrollment rates across indications

Consider individual futility when:

1. **Patient welfare priority**: Maximum protection for individual cohorts
2. **Heterogeneous biology**: Indications likely to have very different outcomes
3. **Sustained enrollment**: All indications expected to enroll at similar rates
4. **Sample size critical**: Minimizing expected enrollment is paramount

## Simulation and Validation

### Simulating Trial Conduct

```{r simulate_example, eval = FALSE}
# Simulate a single trial under the alternative hypothesis
# Assume indications 1 and 3 are active, 2 and 4 are inactive
p_true <- c(0.20, 0.05, 0.35, 0.20)  # True response rates

trial_result <- simulate_basket_trial(
  design = design_heterog,
  p_true = p_true,
  allocation_probs = rep(1/4, 4)  # Equal allocation probability
)

# Examine results
print(trial_result)
```

The simulation shows:
- Whether the trial stopped at Stage I or continued
- Which indications were pruned
- Whether H₀ was rejected (efficacy claimed)
- Detailed enrollment and response data

### Operating Characteristics

To evaluate design performance across scenarios:

```{r operating_characteristics, eval = FALSE}
# Evaluate performance across all possible configurations
oc <- simulate_basket_operating_characteristics(
  design = design_heterog,
  n_sims = 10000,           # 10,000 simulated trials per scenario
  allocation_probs = rep(1/4, 4)
)

# View summary
print(oc)

# Key metrics:
# - Type I error: Should be ≤ 0.05
# - Power by number of active indications
# - Expected number of true/false positives
# - Probability of identifying specific active indications
```

## Effect of Number of Indications

Sample size requirements generally increase with more indications:

```{r k_comparison, eval = FALSE}
# Compare K = 3, 4, 5 indications (homogeneous setting)
results <- data.frame(
  K = c(3, 4, 5),
  S = NA,
  Total_N = NA,
  EN_H0 = NA
)

for (i in 1:3) {
  K_val <- results$K[i]
  design <- two_stage_basket_trial(
    K = K_val,
    p0 = rep(0.05, K_val),
    pa = rep(0.20, K_val),
    alpha = 0.05,
    beta = 0.20,
    homogeneous = TRUE
  )
  results$S[i] <- design$design$S
  results$Total_N[i] <- design$design$total_N
  results$EN_H0[i] <- design$design$EN_H0
}

print(results)
```

## Important Considerations

### Scientific Rationale

**Critical Question**: Do we expect similar treatment effects across indications?

- **Strong rationale**: Same molecular target, similar tumor biology → aggregated approach reasonable
- **Weak rationale**: Different mechanisms, varying baseline rates → consider individual futility

### Enrollment Patterns

**Assess expected enrollment rates:**
- If one indication enrolls at 2 pts/month and others at 8-10 pts/month, aggregated futility occurs much earlier
- If rates are similar, aggregated approach offers less timing advantage

**Minimum sample sizes:**
You can modify the design to require minimum enrollment per indication before futility analysis:
```{r min_sample, eval = FALSE}
# Not implemented in current function, but conceptually:
# Only conduct aggregated analysis after each indication has ≥ n_min patients
```

### Stage I Decision Implications

The aggregated futility decision may:
- **Continue** enrollment in truly inactive indications if fast-enrolling cohorts show responses
- **Terminate** enrollment in truly active slow-enrolling cohorts if fast-enrolling cohorts lack responses

**Mitigation Strategies:**
1. Delay futility analysis until slow-enrolling cohorts reach meaningful size
2. Conduct separate analyses for fast vs. slow enrolling groups
3. Use adaptive allocation to balance enrollment

### Interpretation and Next Steps

**If efficacy is declared:**
- **Do not assume** all pooled indications are active
- **Further investigation** recommended before Phase III:
  - Enroll additional patients in promising indications
  - Consider "2-in-1" designs (seamless Phase II/III)
  - Evaluate biomarkers or response patterns
  
**Individual indication conclusions:**
- Type I and II errors are **not controlled** at the indication level
- Indications included in final pool have higher probability of being active
- Pruned indications may still warrant investigation with larger sample sizes

## Design Optimization

The `two_stage_basket_trial()` function uses numerical optimization to:

1. **Minimize** expected sample size under H₀ (when all indications are inactive)
2. **Subject to** constraints:
   - Global type I error ≤ α (typically 0.05)
   - Expected type II error ≤ β (typically 0.20)

This balances efficiency (fewer patients under null) with power (adequate detection under alternative).

## Comparison to Related Designs

### vs. Simon's Two-Stage Design

| Feature | Simon | Basket (Aggregated) |
|---------|-------|---------------------|
| Indications | 1 | Multiple (K ≥ 2) |
| Hypothesis | Single | Global (any active) |
| Interim | One cohort | Pooled across cohorts |
| Final | One cohort | Pruning + pooling |
| Flexibility | Fixed n1, n2 | Adaptive Stage I allocation |

### vs. Individual Futility Basket Design

| Feature | Individual | Aggregated |
|---------|-----------|------------|
| Stage I | K analyses | 1 pooled analysis |
| Timing | Variable by cohort | Single timepoint |
| Sample size at interim | Pre-specified per cohort | Total only (S) |
| Decision speed | Slowest cohort | Fixed at S patients |
| Expected N under H₀ | Lower | Higher |
| True positive detection | Lower | Higher |

## Example Trial Schemas

### Schema 1: BRAF-Mutant Basket (Homogeneous)

```
Eligibility: BRAF V600E mutation, any solid tumor
Indications: Melanoma, Colorectal, Thyroid, NSCLC
Null ORR: 10% per indication
Target ORR: 30% per indication
```

### Schema 2: MSI-High Basket (Heterogeneous)

```
Eligibility: MSI-H or dMMR, refractory to standard therapy
Indication 1 (Colorectal): p0 = 5%, pa = 20%
Indication 2 (Endometrial): p0 = 10%, pa = 25%  
Indication 3 (Gastric): p0 = 5%, pa = 20%
Indication 4 (Other GI): p0 = 10%, pa = 25%
```

### Schema 3: Rare Mutation Basket (Small K)

```
Eligibility: NTRK fusion, any solid tumor
Indications: Salivary gland, Sarcoma, Other
Null ORR: 5% per indication
Target ORR: 25% per indication
K = 3, balanced enrollment expected
```

## Regulatory Considerations

**FDA Perspective (Draft Guidance on Master Protocols):**
- Basket trials acceptable for exploratory Phase II
- Clear pre-specification of decision rules required
- Global hypotheses appropriate for early development
- Individual cohort confirmation needed for registration

**Key Requirements:**
1. **Prospective definition** of all design parameters
2. **Statistical justification** for pooling approach
3. **Clear decision criteria** at each stage
4. **Follow-up plans** for promising signals

## Session Information

```{r session_info}
sessionInfo()
```

## References
