---
title: "The design grammar"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The design grammar}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

blueprint organizes trial design work around a small set of verbs:

1. **Search or construct** a design: `search_simon_designs()` finds candidates
   meeting your error-rate constraints; `design_simon()` builds a design
   directly from known parameters.
2. **Assume** a data-generating truth: `assume_null()`, `assume_alternative()`,
   or any `assume_response()` scenario you care about.
3. **Simulate and evaluate**: `simulate_trial()` generates tidy trial
   realizations; `evaluate()` returns operating characteristics, exact or
   simulated.
4. **Verify**: `verify()` puts exact and simulated operating characteristics
   side by side, confirming that the design delivers what it promises.

This vignette walks the full workflow with a Simon two-stage design, the
workhorse of single-arm phase II oncology trials.

```{r setup}
library(blueprint)
```

## Search for a design

Suppose a response rate of 20% is uninteresting and 40% would justify further
study. We want one-sided type I error at most 5% and power at least 80%.

```{r search}
candidates <- search_simon_designs(
  p0 = 0.20,
  pa = 0.40,
  alpha = 0.05,
  power = 0.80,
  n_max = 60
)
candidates
```

The search returns one candidate per total sample size, each already meeting
the constraints. Two are flagged: the optimal design minimizes the expected
sample size when the treatment is inactive, and the minimax design minimizes
the worst-case total. Picking one gives a design object.

```{r pick}
design <- pick_design(candidates, "optimal")
design
```

If the trade-off matters for your setting (the minimax design here caps
enrollment at 33 instead of 43, at the cost of a larger expected sample
size), pick the other flag or construct any row directly with
`design_simon()`.

## Reproduce a design from a protocol

Search is not the only door in. When a protocol already specifies the design,
construct it directly and interrogate it:

```{r construct}
protocol_design <- design_simon(
  n1 = 18, r1 = 4, n = 33, r = 10,
  p0 = 0.20, pa = 0.40
)
evaluate(protocol_design)
```

## Verify by simulation

Exact calculations say what the design should do. `verify()` checks that
simulated trials agree:

```{r verify}
verify(design, n_sims = 10000, seed = 2026)
```

## Evaluate scenarios the design was not built for

Designs get built for two hypotheses, but trials happen in between. Scenario
objects make it easy to ask how the design behaves at any truth:

```{r scenarios}
ocs <- evaluate(design, under = list(
  "p = 0.20 (null)" = assume_response(0.20),
  "p = 0.25" = assume_response(0.25),
  "p = 0.30" = assume_response(0.30),
  "p = 0.35" = assume_response(0.35),
  "p = 0.40 (alternative)" = assume_response(0.40)
))
ocs
```

Because the result is a tibble, it drops straight into a report table:

```{r gt, eval = requireNamespace("gt", quietly = TRUE)}
ocs |>
  gt::gt() |>
  gt::cols_label(
    scenario = "Scenario",
    p_true = "True rate",
    prob_success = "P(reject null)",
    prob_early_stop = "P(stop at stage 1)",
    expected_n = "E[N]"
  ) |>
  gt::cols_hide("method") |>
  gt::fmt_number(columns = c("prob_success", "prob_early_stop"), decimals = 3) |>
  gt::fmt_number(columns = "expected_n", decimals = 1)
```

## Simulate individual trials

`simulate_trial()` returns one row per simulated trial, which is useful when
you want the full distribution rather than its summaries:

```{r simulate}
sims <- simulate_trial(
  design,
  under = assume_response(0.30),
  n_sims = 5000,
  seed = 2026
)
sims

table(sims$n_enrolled) / nrow(sims)
```

## Draft the protocol paragraph

`draft_protocol_text()` writes the sample size justification in the phrasing
conventional for protocols. It returns a plain string, so it works inline in
Quarto or R Markdown:

```{r protocol, results = "asis"}
cat(draft_protocol_text(design))
```

## Tidy summaries

`tidy()` and `glance()` follow broom conventions: `tidy()` gives the
operating characteristics under the design's own hypotheses, and `glance()`
gives a one-row design summary for tabulating several designs together.

```{r tidy}
tidy(design)
glance(design)
```

## References

Simon, R. (1989). Optimal two-stage designs for phase II clinical trials.
*Controlled Clinical Trials*, 10(1), 1-10.
https://doi.org/10.1016/0197-2456(89)90015-9
