---
title: "Simple Basket Trial Design"
author: "Travis Gerke"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simple Basket Trial Design}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Overview

The `simple_basket_trial()` function implements the adaptive two-stage basket trial design proposed by @cunanan2017. This design is **simpler and more efficient** than conducting independent trials in each basket when the targeted therapy works in most or all tumor types.

### Key Features

- **Adaptive design**: Interim heterogeneity assessment determines analysis strategy
- **Two paths**:
  - **Heterogeneous**: Analyze baskets separately if response rates differ
  - **Homogeneous**: Pool all baskets if response rates similar
- **Frequentist framework**: Fisher's exact test and binomial tests (no Bayesian MCMC)
- **Efficiency gains**: Up to 36% fewer patients when drug works in all baskets
- **FWER control**: Family-wise error rate controlled at target level

### When to Use This Design

✅ **Use when:**
- Testing a targeted therapy across multiple tumor types with shared molecular alteration
- Drug is expected to work in most/all baskets (homogeneous efficacy)
- Want simpler alternative to Bayesian basket designs
- Need frequentist framework with clear Type I error control

❌ **Don't use when:**
- Drug only expected to work in 1-2 baskets (use independent trials or Bayesian borrowing)
- Strong FWER control required across all scenarios (this design has weak control)
- Very heterogeneous effects expected across baskets

## Methodology

### Design Structure

The simple basket trial proceeds in two stages with an interim analysis:

**Stage 1**: Accrue N₁ patients total across K baskets

- Minimum 3 patients per basket (default)
- Maximum 10 patients per basket (default)
- Example: N₁=35 for K=5 baskets ≈ 7 per basket

**Interim Analysis**: Test of Homogeneity (TOH)

- Fisher's exact test on K×2 contingency table (responders vs non-responders)
- Critical value γ (tuning parameter between 0 and 1)
- **If p-value ≤ γ**: Pursue **heterogeneous path** (baskets appear different)
- **If p-value > γ**: Pursue **homogeneous path** (baskets appear similar)

**Stage 2 - Heterogeneous Path**:

For each basket k:

1. **Futility rule**: If < r_s responders in stage 1, stop basket k
2. Otherwise, accrue n₂ₖ additional patients to basket k
3. **Final analysis**: Binomial test at level αS/K* (Bonferroni correction)
   - K* = number of continuing baskets
4. Reject H₀ for basket k if p-value ≤ αS/K*

**Stage 2 - Homogeneous Path**:

1. **Futility rule**: If < r_c total responders in stage 1, stop all baskets
2. Otherwise, accrue N₂ total additional patients (pooled across baskets)
3. **Final analysis**: One-sample binomial test at level αC
4. Reject H₀ for **all baskets** or **none**

### Notation

| Symbol | Description |
|--------|-------------|
| K | Number of baskets |
| θ₀ | Null response rate |
| θₐ | Alternative (active) response rate |
| A | Number of truly active baskets (0 to K) |
| N₁ | Total stage 1 sample size across all baskets |
| N₂ | Total stage 2 sample size (homogeneous path) |
| n₂ₖ | Stage 2 sample size per basket (heterogeneous path) |
| r_s | Minimum responders per basket to continue (heterogeneous) |
| r_c | Minimum total responders to continue (homogeneous) |
| γ | Heterogeneity test critical value (0-1) |
| αS | Significance level for separate analyses (heterogeneous) |
| αC | Significance level for combined analysis (homogeneous) |
| ε | Target family-wise error rate (FWER) |
| 1-β | Target power |

### Operating Characteristics

For each scenario (A = number of active baskets):

- **FWER**: P(reject ≥1 basket | all null) — only relevant when A=0
- **Marginal Power**: P(reject basket k | basket k active)
- **EN**: Expected total sample size
- **ET**: Expected trial duration (months)
- **Sensitivity**: P(correctly identify active basket)
- **Specificity**: P(correctly identify inactive basket)

## Example: Replicating Cunanan et al. (2017) Table 2

The paper presents a K=5 basket design with θ₀=0.15 and θₐ=0.45. Let's replicate their results.

### Design Optimization

```{r design, eval=FALSE}
library(blueprint)

# Replicate Table 2: K=5, θ0=0.15, θa=0.45
design <- simple_basket_trial(
  K = 5,
  p0 = 0.15,
  pa = 0.45,
  N1 = 35,              # Total stage 1 sample size
  N2 = 20,              # Total stage 2 (homogeneous path)
  r_s = 1,              # Min responders per basket to continue
  r_c = 5,              # Min total responders to continue
  target_fwer = 0.05,   # 5% family-wise error rate
  target_power = 0.80,  # 80% power at calibration scenario
  min_power = 0.70,     # 70% minimum power when A=1
  accrual_rate = 2,     # 2 patients/month per basket
  calibrate_at_A = 2,   # Calibrate power when A=2 active
  # Narrow grid for faster computation
  n2k_range = 13:17,
  gamma_range = seq(0.4, 0.6, by = 0.05),
  alpha_s_range = seq(0.05, 0.09, by = 0.01),
  alpha_c_range = seq(0.04, 0.06, by = 0.01),
  n_sims = 1000,
  seed = 4206
)

# View design
print(design)
```

**Expected optimal parameters** (from Cunanan et al. Table 2):

- n₂ₖ = 15 patients
- γ = 0.52
- αS = 0.07
- αC = 0.05

### Operating Characteristics

```{r summary, eval=FALSE}
# Detailed operating characteristics
summary(design)
```

**Expected results** (from Table 2):

| Scenario | FWER/Power | EN | ET (months) |
|----------|------------|-----|-------------|
| A=0 (all null) | FWER=5% | 58 | 7.0 |
| A=1 | Power=70% | 74 | 9.5 |
| A=2 (calibration) | Power=80% | 83 | 10.4 |
| A=3 | Power=84% | 86 | 10.5 |
| A=4 | Power=86% | 88 | 10.2 |
| A=5 (all active) | Power=88% | 78 | 8.3 |

### Key Insights

1. **FWER Control**: The design maintains FWER ≤ 5% when all baskets are null

2. **Calibrated Power**: Achieves 80% power when 2 baskets are active (as designed)

3. **Minimum Power**: Maintains ≥70% power even when only 1 basket is active

4. **Efficiency Gains**: 
   - A=5: EN=78 vs 121 for independent designs (36% reduction!)
   - A=0: EN=58 (early futility stopping)
   
5. **Power Increases with A**: Due to pooling in homogeneous path

6. **Trade-off**: Specificity decreases as A increases (risk of false positives due to pooling)

## Understanding the Two Paths

### Heterogeneous Path Example

Suppose after stage 1:
- Basket 1: 4/7 responses (57%)
- Basket 2: 3/7 responses (43%)
- Basket 3: 0/7 responses (0%)
- Basket 4: 1/7 responses (14%)
- Basket 5: 1/7 responses (14%)

Fisher's exact test p-value = 0.03 < γ=0.52 → **Heterogeneous**

Continue baskets 1, 2, 4, 5 (≥1 responder). Stop basket 3 for futility.

Accrue n₂ₖ=15 more patients per continuing basket.

Final analysis: Test each basket separately with α = 0.07/4 = 0.0175

### Homogeneous Path Example

Suppose after stage 1:
- Basket 1: 3/7 responses (43%)
- Basket 2: 3/7 responses (43%)
- Basket 3: 2/7 responses (29%)
- Basket 4: 2/7 responses (29%)
- Basket 5: 2/7 responses (29%)

Fisher's exact test p-value = 0.95 > γ=0.52 → **Homogeneous**

Total responses = 12 ≥ r_c=5 → Continue to stage 2

Accrue N₂=20 total additional patients (pooled).

Final analysis: One-sample test with α = 0.05

If significant → Declare **all baskets active**  
If not significant → Declare **all baskets inactive**

## Comparison to Other Designs

### vs. Independent Simon Two-Stage Designs

**Advantages of Simple Basket:**
- ✅ 36% fewer patients when drug works in all baskets (78 vs 121)
- ✅ Faster trial completion when homogeneous
- ✅ Higher power when A ≥ 3

**Disadvantages:**
- ❌ Lower power when A=1 (70% vs 79%)
- ❌ Slightly more patients when A=1 (74 vs 69)
- ❌ Risk of false positives in inactive baskets (specificity ↓ as A ↑)

### vs. Bayesian Basket Designs (e.g., Berry et al.)

**Advantages of Simple Basket:**
- ✅ Simpler implementation (no MCMC)
- ✅ Faster computation
- ✅ Frequentist framework (clearer Type I error interpretation)
- ✅ No prior specification required

**Disadvantages:**
- ❌ Less flexible borrowing across baskets
- ❌ Weak FWER control (only at A=0)
- ❌ All-or-none decision in homogeneous path

## Practical Considerations

### Choosing N₁, N₂, r_s, r_c

The paper recommends:

1. **N₁**: Choose based on budget/timeline
   - Default: K × (first stage of Simon optimal design)
   - Example: For K=5, N₁ = 35

2. **r_s**: Usually 1 (continue if any response observed)
   - Being conservative avoids missing active baskets

3. **r_c**: Usually K (≈1 per basket)
   - For K=5, r_c = 5 is reasonable

4. **N₂**: Smaller than N₁ (pooling provides power)
   - Should be in range of Simon optimal stage 2
   - Example: N₂ = 20 for K=5

### Unequal Accrual Rates

If baskets accrue at different rates, specify `accrual_rate` as vector:

```{r unequal, eval=FALSE}
design_unequal <- simple_basket_trial(
  K = 5,
  p0 = 0.15,
  pa = 0.45,
  N1 = 35,
  N2 = 20,
  r_s = 1,
  r_c = 5,
  accrual_rate = c(3, 2, 1, 1, 1),  # Fast accrual in baskets 1-2
  # ... other parameters
)
```

**Impact**: 
- Trial duration increases substantially
- Power may increase if active baskets accrue fast
- Power may decrease if active baskets accrue slow

### Calibration Strategy

For **K=5 baskets**: Use `calibrate_at_A = 2`

For **K≥10 baskets**: Use `calibrate_at_A = 3`

**Why?** The calibration scenario should be "middle ground":
- Not too pessimistic (A=1)
- Not too optimistic (A=K)
- Provides robust operating characteristics across scenarios

### Computational Performance

Grid search can be slow for large grids. To speed up:

1. **Narrow the search ranges** around expected optimal values
2. **Use parallel processing**: `parallel = TRUE, n_cores = 8`
3. **Reduce n_sims** initially (use 100-200 for exploration, 1000+ for final)
4. **Two-phase search**:
   - Phase 1: Coarse grid with n_sims=200
   - Phase 2: Fine grid around optimal with n_sims=1000

## Simulation-Based Validation

To validate a specific design, compute operating characteristics directly:

```{r validate, eval=FALSE}
# Compute OCs for specific design parameters
oc_A0 <- simulate_simple_operating_characteristics(
  K = 5,
  A = 0,  # All baskets null
  p0 = 0.15,
  pa = 0.45,
  N1 = 35,
  N2 = 20,
  n2k = 15,
  r_s = 1,
  r_c = 5,
  gamma = 0.52,
  alpha_s = 0.07,
  alpha_c = 0.05,
  accrual_rate = rep(2, 5),
  n_sims = 5000,
  seed = 123
)

cat("FWER:", oc_A0$fwer, "\n")
cat("Expected N:", oc_A0$EN, "\n")
cat("Expected duration:", oc_A0$ET, "months\n")

# A=2 scenario
oc_A2 <- simulate_simple_operating_characteristics(
  K = 5, A = 2,  # First 2 baskets active
  # ... same parameters ...
  n_sims = 5000
)

cat("Power (Basket 1):", oc_A2$marginal_power[1], "\n")
cat("Power (Basket 2):", oc_A2$marginal_power[2], "\n")
cat("Expected N:", oc_A2$EN, "\n")
```

## Reporting Guidelines

When reporting results from a simple basket trial, include:

1. **Design parameters**: K, N₁, N₂, n₂ₖ, r_s, r_c, γ, αS, αC
2. **Operating characteristics** for all scenarios (A=0 to K)
3. **Calibration strategy** and justification
4. **Actual trial results**:
   - Stage 1 sample sizes and responses per basket
   - Heterogeneity test p-value and path selected
   - Stage 2 sample sizes and responses
   - Final decisions per basket
   - Total sample size and duration
5. **Comparison to planned OCs**

## References

<div id="refs"></div>

## Session Info

```{r session}
sessionInfo()
```
