---
title: "Single-Stage Design: Simulations and Operating Characteristics"
author: "Travis Gerke"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Single-Stage Design: Simulations and Operating Characteristics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
library(blueprint)
```

## Introduction

This article demonstrates the complete workflow for designing and validating single-stage single-arm phase II trials with time-to-event endpoints using the `single_stage_single_arm_tte()` function. We'll:

1. Design a single-stage trial using the Wu (2015) methodology
2. Simulate individual trial realizations to understand trial behavior
3. Estimate operating characteristics through repeated simulations
4. Compare two-sided vs one-sided tests
5. Examine the impact of different follow-up times

## Background

The `single_stage_single_arm_tte()` function implements the methodology from Wu (2015) for single-stage designs with time-to-event endpoints. Key features include:

- **Two-sided tests** (default): Uses critical value $z_{\alpha/2}$ for testing in both directions
- **One-sided tests**: Uses critical value $z_\alpha$ for one-directional testing (smaller sample size)
- **Parametric distributions**: Supports Weibull, Log-normal, Gamma, and Log-logistic
- **Follow-up models**: Restricted (everyone followed exactly `tf` time units) or unrestricted (staggered follow-up based on enrollment)

## Example 1: Basic Single-Stage Design

Let's start with a simple exponential survival distribution (Weibull with shape = 1).

**Scenario:**
- Null hypothesis: Median survival = 0.5 years (poor prognosis)
- Alternative hypothesis: Median survival = 0.75 years (improved treatment)
- Hazard ratio: 0.667 (33% reduction in hazard)
- Follow-up time: 1 year
- Accrual rate: 32.4 patients/year
- Two-sided test with α = 0.05, power = 90%

```{r design1}
# Design the trial
design_exponential <- single_stage_single_arm_tte(
  shape = 1.0,        # Exponential distribution
  S0 = 0.5,           # 50% survival at x0
  x0 = 0.5,           # Median survival under H0
  hr = 0.667,         # 33% hazard reduction
  tf = 1.0,           # 1 year follow-up
  rate = 32.4,        # Accrual rate
  alpha = 0.05,       # Two-sided 5% Type I error
  beta = 0.10,        # 90% power
  two_sided = TRUE,   # Two-sided test
  dist = "WB",        # Weibull distribution
  restricted = FALSE  # Unrestricted follow-up
)

print(design_exponential)
```

The design requires **n = `r design_exponential$single_stage$n`** patients with an accrual time of **ta = `r round(design_exponential$single_stage$ta, 2)`** years. The critical value is **c = `r round(design_exponential$single_stage$c, 3)`**, and we expect approximately **E = `r round(design_exponential$single_stage$E, 1)`** events.

## Example 2: Simulating Individual Trials

Now let's simulate what a single trial realization might look like under both the null and alternative hypotheses.

```{r simulate_single_h0}
# Simulate one trial under H0 (median = 0.5)
set.seed(2024)

# Calculate scale parameter for Weibull distribution
# For Weibull: median = scale * (log(2))^(1/shape)
# So: scale = median / (log(2))^(1/shape)
shape_h0 <- 1.0
median_h0 <- 0.5
scale_h0 <- median_h0 / (log(2))^(1/shape_h0)

trial_h0 <- simulate_trial(
  design = design_exponential,
  shape = shape_h0,
  scale = scale_h0
)

cat("Trial under H0:\n",
    "  Test statistic Z = ", round(trial_h0$final_stat, 3), "\n",
    "  Critical value c = ", round(design_exponential$single_stage$c, 3), "\n",
    "  Decision: ", ifelse(trial_h0$reject_h0, "Reject H0", "Fail to reject H0"), "\n",
    "  Number of events: ", trial_h0$final_events, "\n", sep = "")
```

```{r simulate_single_h1}
# Simulate one trial under H1 (median = 0.75)
set.seed(2025)

# Calculate scale parameter for H1
median_h1 <- 0.75
scale_h1 <- median_h1 / (log(2))^(1/shape_h0)

trial_h1 <- simulate_trial(
  design = design_exponential,
  shape = shape_h0,
  scale = scale_h1
)

cat("Trial under H1:\n",
    "  Test statistic Z = ", round(trial_h1$final_stat, 3), "\n",
    "  Critical value c = ", round(design_exponential$single_stage$c, 3), "\n",
    "  Decision: ", ifelse(trial_h1$reject_h0, "Reject H0", "Fail to reject H0"), "\n",
    "  Number of events: ", trial_h1$final_events, "\n", sep = "")
```

### Understanding the Output

For single-stage trials, the simulation output includes:

- **final_stat**: The log-rank test statistic Z
- **reject_h0**: TRUE if null hypothesis is rejected (Z > c)
- **stopped_early**: FALSE for single-stage designs (no interim analysis)
- **enrolled_n**: Total number of patients enrolled
- **final_events**: Number of events observed
- **final_data**: Data frame with patient-level data (enrollment times, event times, censoring)
- **final_time**: Calendar time of final analysis

## Example 3: Operating Characteristics via Simulation

To verify the design properties, we simulate many trials and estimate the operating characteristics.

```{r operating_characteristics}
# Estimate operating characteristics with 1000 simulations
set.seed(123)

# Calculate scale parameters
# H0: median = 0.5
# H1: median = 0.75
shape_param <- 1.0
scale_h0 <- 0.5 / (log(2))^(1/shape_param)
scale_h1 <- 0.75 / (log(2))^(1/shape_param)

oc_results <- simulate_operating_characteristics(
  design = design_exponential,
  shape = shape_param,
  scale_h0 = scale_h0,
  scale_h1 = scale_h1,
  n_sims = 1000
)

cat("\nOperating Characteristics (1000 simulations):\n",
    "  Type I Error Rate (α): ", round(oc_results$type1_error, 4), "\n",
    "  Power (1-β):           ", round(oc_results$power, 4), "\n",
    "  Expected N under H0:   ", round(oc_results$expected_n_h0, 1), "\n",
    "  Expected N under H1:   ", round(oc_results$expected_n_h1, 1), "\n",
    "  Avg events under H0:   ", round(oc_results$avg_final_events_h0, 1), "\n",
    "  Avg events under H1:   ", round(oc_results$avg_final_events_h1, 1), "\n", sep = "")
```

**Interpretation:**

- The empirical **Type I error rate** should be close to α = 0.05
- The empirical **power** should be close to 1-β = 0.90
- For single-stage designs, **expected N** equals the planned sample size (no early stopping)
- The average number of events reflects censoring patterns

### Distribution of Test Statistics

Let's visualize the distribution of test statistics under H0 and H1:

```{r plot_test_stats, fig.width=8, fig.height=5, fig.alt="Two histograms showing the distribution of test statistics. Left panel shows test statistics under H0 following a standard normal distribution with critical value marked. Right panel shows test statistics under H1 shifted right with most values exceeding the critical value."}
# Extract test statistics from individual simulations
test_stats_h0 <- sapply(oc_results$results_h0, function(x) x$final_stat)
test_stats_h1 <- sapply(oc_results$results_h1, function(x) x$final_stat)

# Create histogram
par(mfrow = c(1, 2))
hist(test_stats_h0, breaks = 30, main = "Test Statistics under H0",
     xlab = "Z statistic", col = "lightblue", probability = TRUE)
abline(v = design_exponential$single_stage$c, col = "red", lwd = 2, lty = 2)
curve(dnorm(x, 0, 1), add = TRUE, col = "blue", lwd = 2)
legend("topright", c("Simulated", "N(0,1)", "Critical value"),
       col = c("lightblue", "blue", "red"), lty = c(1, 1, 2), lwd = c(10, 2, 2))

hist(test_stats_h1, breaks = 30, main = "Test Statistics under H1",
     xlab = "Z statistic", col = "lightgreen", probability = TRUE)
abline(v = design_exponential$single_stage$c, col = "red", lwd = 2, lty = 2)
legend("topright", c("Simulated", "Critical value"),
       col = c("lightgreen", "red"), lty = c(1, 2), lwd = c(10, 2))
```

Under H0, the test statistic should follow a standard normal distribution. Under H1, it should be shifted to the right, with approximately 90% of values exceeding the critical value.

## Example 4: Two-Sided vs One-Sided Tests

A key feature of the single-stage design is the option for two-sided vs one-sided tests. Let's compare:

```{r compare_sided}
# Two-sided test (already computed above)
cat("Two-sided test (α = 0.05):\n",
    "  Sample size:       ", design_exponential$single_stage$n, "\n",
    "  Critical value:    ", round(design_exponential$single_stage$c, 3), "\n",
    "  Expected events:   ", round(design_exponential$single_stage$E, 1), "\n\n", sep = "")

# One-sided test
design_one_sided <- single_stage_single_arm_tte(
  shape = 1.0,
  S0 = 0.5,
  x0 = 0.5,
  hr = 0.667,
  tf = 1.0,
  rate = 32.4,
  alpha = 0.05,
  beta = 0.10,
  two_sided = FALSE,  # One-sided test
  dist = "WB",
  restricted = FALSE
)

cat("One-sided test (α = 0.05):\n",
    "  Sample size:       ", design_one_sided$single_stage$n, "\n",
    "  Critical value:    ", round(design_one_sided$single_stage$c, 3), "\n",
    "  Expected events:   ", round(design_one_sided$single_stage$E, 1), "\n", sep = "")
```

**Key differences:**

- Two-sided: c = `r round(design_exponential$single_stage$c, 3)` (qnorm(0.975))
- One-sided: c = `r round(design_one_sided$single_stage$c, 3)` (qnorm(0.95))
- One-sided designs require **fewer patients** due to lower critical value

## Example 5: Impact of Follow-Up Time

Follow-up time (`tf`) significantly affects sample size. Let's examine this:

```{r followup_comparison}
# Compare different follow-up times
followup_times <- c(0.5, 1.0, 1.5, 2.0)
results <- data.frame(
  tf = followup_times,
  n = NA,
  ta = NA,
  E = NA,
  P1 = NA
)

for (i in seq_along(followup_times)) {
  design_temp <- single_stage_single_arm_tte(
    shape = 1.0,
    S0 = 0.5,
    x0 = 0.5,
    hr = 0.667,
    tf = followup_times[i],
    rate = 32.4,
    alpha = 0.05,
    beta = 0.10,
    two_sided = TRUE,
    dist = "WB",
    restricted = FALSE
  )
  
  results$n[i] <- design_temp$single_stage$n
  results$ta[i] <- round(design_temp$single_stage$ta, 2)
  results$E[i] <- round(design_temp$single_stage$E, 1)
  results$P1[i] <- round(design_temp$single_stage$P1, 4)
}

print(results)
```

**Observations:**

- **Longer follow-up** → more events per patient → **smaller sample size**
- **P1** (probability of event) increases with follow-up time
- Trade-off: Fewer patients vs longer study duration

### Visualize the Relationship

```{r plot_followup, fig.width=8, fig.height=5, fig.alt="Two line plots showing the relationship between follow-up time and design parameters. Left panel shows sample size decreasing as follow-up time increases. Right panel shows expected events increasing with follow-up time."}
par(mfrow = c(1, 2))

# Sample size vs follow-up
plot(results$tf, results$n, type = "b", pch = 19, col = "blue",
     xlab = "Follow-up time (years)", ylab = "Sample size",
     main = "Sample Size vs Follow-up Time")
grid()

# Expected events vs follow-up
plot(results$tf, results$E, type = "b", pch = 19, col = "darkgreen",
     xlab = "Follow-up time (years)", ylab = "Expected events",
     main = "Expected Events vs Follow-up Time")
grid()
```

## Example 6: Non-Exponential Distributions

The methodology supports various parametric distributions. Let's compare designs with different shape parameters:

```{r shapes_comparison}
# Weibull with different shapes
shapes <- c(1.0, 1.25, 1.5, 2.0)
shape_results <- data.frame(
  shape = shapes,
  distribution = c("Exponential", "Increasing hazard", 
                   "Increasing hazard", "Rayleigh"),
  n = NA,
  E = NA,
  P1 = NA
)

for (i in seq_along(shapes)) {
  design_temp <- single_stage_single_arm_tte(
    shape = shapes[i],
    S0 = 0.5,
    x0 = 0.5,
    hr = 0.667,
    tf = 1.0,
    rate = 32.4,
    alpha = 0.05,
    beta = 0.10,
    two_sided = TRUE,
    dist = "WB",
    restricted = FALSE
  )
  
  shape_results$n[i] <- design_temp$single_stage$n
  shape_results$E[i] <- round(design_temp$single_stage$E, 1)
  shape_results$P1[i] <- round(design_temp$single_stage$P1, 4)
}

print(shape_results)
```

**Key insights:**

- **Shape = 1.0**: Exponential (constant hazard)
- **Shape > 1.0**: Increasing hazard over time
- Higher shape parameters can lead to different sample size requirements depending on the timing of events

## Example 7: Verifying Design Under Alternative Scenarios

Let's verify that the design maintains correct Type I error under the null and achieves target power under the alternative:

```{r verify_design, eval=FALSE}
# WARNING: This takes ~10 minutes to run
# Set eval=TRUE if you want to run it

# More simulations for precise estimates
set.seed(456)
oc_precise <- simulate_operating_characteristics(
  design = design_exponential,
  shape = shape_param,
  scale_h0 = scale_h0,
  scale_h1 = scale_h1,
  n_sims = 5000  # More simulations for precision
)

cat("\nPrecise Operating Characteristics (5000 simulations):\n",
    "  Type I Error Rate: ", round(oc_precise$type1_error, 4), 
    " ± ", round(sqrt(oc_precise$type1_error * (1 - oc_precise$type1_error) / 5000), 4), "\n",
    "  Power:             ", round(oc_precise$power, 4), 
    " ± ", round(sqrt(oc_precise$power * (1 - oc_precise$power) / 5000), 4), "\n", sep = "")
```

## Summary

This article demonstrated:

1. ✅ **Designing** single-stage trials with `single_stage_single_arm_tte()`
2. ✅ **Simulating** individual trial realizations with `simulate_trial()`
3. ✅ **Validating** operating characteristics with `simulate_operating_characteristics()`
4. ✅ **Comparing** two-sided vs one-sided tests
5. ✅ **Optimizing** follow-up time and understanding trade-offs
6. ✅ **Exploring** different distributional assumptions

### Recommended Workflow

For your own single-stage design:

```r
library(blueprint)

# 1. Design the trial
my_design <- single_stage_single_arm_tte(
  shape = <your_shape>,
  S0 = <survival_prob>,
  x0 = <time_point>,
  hr = <hazard_ratio>,
  tf = <followup_time>,
  rate = <accrual_rate>,
  alpha = 0.05,
  beta = 0.20,
  two_sided = TRUE,  # or FALSE
  dist = "WB",       # or "LN", "GM", "LG"
  restricted = FALSE
)

# 2. Examine the design
print(my_design)

# 3. Calculate scale parameters for simulation
# For Weibull: median = scale * (log(2))^(1/shape)
# So: scale = median / (log(2))^(1/shape)
shape_param <- <your_shape>
median_h0 <- <null_median>
median_h1 <- <alt_median>
scale_h0 <- median_h0 / (log(2))^(1/shape_param)
scale_h1 <- median_h1 / (log(2))^(1/shape_param)

# 4. Simulate individual trials to understand behavior
trial_example <- simulate_trial(
  design = my_design,
  shape = shape_param,
  scale = scale_h1  # Use scale_h0 for null, scale_h1 for alternative
)

# 5. Validate operating characteristics
oc <- simulate_operating_characteristics(
  design = my_design,
  shape = shape_param,
  scale_h0 = scale_h0,
  scale_h1 = scale_h1,
  n_sims = 1000  # Increase for precision
)

# 6. Check Type I error and power
cat("Type I error: ", round(oc$type1_error, 3), "\n",
    "Power:        ", round(oc$power, 3), "\n", sep = "")
```

## References

Wu, J. (2015). Sample size calculation for the one-sample log-rank test. 
*Pharmaceutical Statistics*, 14(1), 26-33. https://doi.org/10.1002/pst.1654
